Elite individuals, insitutions, and economic growth accounting

### 2024-04-07 presentation/main.tex
should be done

\begin{frame}
    \frametitle{Research Question}
    \begin{block}{What is the relationship between elite students, academic institutions, and economic growth?}
        \begin{itemize}
            \item Are there outsized returns to economic growth when there are more academic elites?
            \item How can we measure academic competition and education quality beyond means?
        \end{itemize}
    \end{block}
    \begin{exampleblock}{Method}
        Through this paper, I investigate the relationship between the number of top-ranked universities, share of top math students,
        IMO scores, and GDP per capita growth to analyze these questions.
    \end{exampleblock}
\end{frame}

\begin{frame}
\frametitle{Data Sources}
\begin{itemize}
    \item World Bank: World Development Indicators
    \item International Math Olympiad
    \item ARWU University Rankings
    \item PISA Math Scores
    \item Economist Democracy Scores
\end{itemize}
\end{frame}

\begin{frame}{World Bank Data}
    The World Bank collects and publishes development indicators for most countries/economies in the world. In this paper, I use the following variables:

@@

\begin{frame}{Imputing nulls}{Using XGBoost for better predictions}
    \begin{block}{Data is not missing at random}
        School completion data is only available for a maximum of 80 countries per year and has high variance in this availablility.
        This is the most limiting factor in the analysis.
    \end{block}

@@

\begin{frame}{IMO Scores}{The IMO is an international mathematics contest for high school students
    }
    \begin{columns}
        \column{0.5\textwidth}
            \centering
            \begin{table}
            \caption{Top 10 countries by IMO score.}
            \resizebox{\textwidth}{!}{
            \begin{tabular}{lrrr}
                \toprule
                 & IMO Score & GDPpc & GDPpc Growth \\
                Region &  &  &  \\
                \midrule
                KOR & 10.2026 & 26060.9203 & 293.8552 \\
                CHN & 9.8296 & 6424.4972 & 786.8171 \\
                USA & 9.6196 & 54346.5078 & 125.7613 \\
                RUS & 9.0800 & 10432.8366 & 277.0704 \\
                SGP & 9.0642 & 51652.4651 & 348.0543 \\
                BGR & 8.8821 & 7556.3933 & 417.7564 \\
                ROU & 8.7367 & 9432.6698 & 439.7522 \\
                HUN & 8.7153 & 13979.7009 & 262.9650 \\
                VNM & 8.5029 & 2159.2583 & 527.6404 \\
                UKR & 8.3941 & 3079.8378 & 131.9303 \\
                \bottomrule
            \end{tabular}}
        \end{table}
        \column{0.5\textwidth}
            \begin{itemize}
                \item Scores collected from 2003 to 2022
                \item Scores are transformed by $t(s) = \frac{s}{\log P}$ where $s$ is the country's raw score and $P$ is population.
                \begin{itemize}
                    \item Team size of 6 means that larger countries have an advantage due to ``genius odds''
                    \item Score is capped, so dividing by $\log P$ will correct for theoretical ceiling of performance
                \end{itemize}
            \end{itemize}
        \end{columns}
\end{frame}

\begin{frame}{ARWU Rankings}{Description and Usage}
    ARWU (Academic Ranking of World Universities) is a set of university rankings based primarily on research output.
    \begin{itemize}
        \item Rankings are produced annually and are available from 2003
        \item From 2003 to 2016, 500 top universities were ranked; after 2017, 1000 were ranked.
    \end{itemize}

@@

\begin{frame}{EIU Democracy Index}
    \begin{columns}
        \column{0.4\textwidth}
        Index measuring the quality of democracy around the world published by the Economist Intelligence Unit.
        \begin{itemize}
            \item Published from 2006 every 2 years until 2010, annually afterwards (use XGBoost to impute missing)
            \item 0-10 scale, where 10 is democracy and 0 is autocracy.
        \end{itemize}

@@

\begin{frame}{Summary Statistics}
    \begin{table}
        \caption{Summary Statistics}
        \resizebox{\linewidth}{!} {
            \begin{tabular}{lrrrrr}
                & Count & Mean & Std & Min & Max \\
                \\[-1.8ex]\hline
                \hline \\[-1.8ex]
               GDP per capita & 441.00 & 28463.99 & 25255.10 & 543.11 & 149461.79 \\
               GDP per capita growth (bps) & 440.00 & 169.06 & 452.68 & -2292.68 & 3303.05 \\
               EIU Democracy Index & 441.00 & 7.25 & 1.77 & 1.93 & 9.93 \\
               Primary Completion & 441.00 & 90.00 & 10.25 & 51.35 & 101.95 \\
               Lower Sec. Completion & 441.00 & 77.89 & 17.14 & 29.21 & 101.97 \\
               Upper Sec. Completion & 441.00 & 62.20 & 18.48 & 18.44 & 97.40 \\
               Population & 441.00 & 35318600.19 & 59445620.16 & 34000.00 & 333287557.00 \\
               ARWU per million pop ranked & 441.00 & 0.26 & 0.33 & 0.00 & 1.54 \\
               IMO score per log pop & 441.00 & 4.72 & 3.49 & 0.00 & 11.79 \\
               PISA math in global 1\% & 441.00 & 0.94 & 1.46 & 0.00 & 14.64 \\
               PISA math & 441.00 & 461.93 & 56.23 & 315.96 & 574.66 \\
               \end{tabular}
        }
    \end{table}
    \begin{table}
        \tiny
        \caption{Pre-imputation Statistics}
        \begin{tabular}{lrrrrr}
         & Count & Mean & Std & Min & Max \\
         \\[-1.8ex]\hline
         \hline \\[-1.8ex]
        EIU Democracy Index & 323.00 & 7.09 & 1.78 & 1.93 & 9.93 \\
        Primary Completion & 157.00 & 90.29 & 10.31 & 51.35 & 100.00 \\
        Lower Sec. Completion & 179.00 & 76.63 & 20.03 & 29.21 & 99.98 \\
        Upper Sec. Completion & 198.00 & 62.32 & 20.58 & 18.44 & 97.40 \\
        \end{tabular}
        \end{table}
\end{frame}

@@

\begin{frame}{Model specification}
    Let elite indicators be: $math99, ARWU, IMO$.
    For the sake of concision, let $E_{i,t}$ be a 1 by 4 matrix defined as:
    \[E_{i,t} = 
    \begin{bmatrix}
        math99_{i, t} & ARWU_{i, t} & ARWU_{i, t} \times GDPpc_{i, t} & IMO_{i, t}
    \end{bmatrix}
    \] for country/region $i$ in year $t$.
    Let the coefficients of $E_{i, t}$ be a 4 by 1 matrix called $\lambda$. For control variables, $I_{i, t}$ is the matrix of variables and $\alpha$ is coefficients.
    \begin{equation}
        Y_{i, t} = \beta_0 + \lambda E_{i, t} + \delta GDPpc \times E_{i,t} + \alpha I_{i, t} + T_t + C_i + \epsilon_{i, t}
    \end{equation}
    where $T,C$ represent time and entity dummies respectively and $Y_{i,t}$ be the GDP per capita growth in basis points for a country/region $i$ and year $t$.
\end{frame}

\begin{frame}{Relationships}
    \centering
    \includegraphics[width=\textwidth]{../charts/relationships.png}
\end{frame}

\begin{frame}{PISA panel regression}{Highly dependent on model specification due to rich country bias and limited variation}
    \begin{table}[!htbp] \centering
        \resizebox{\linewidth}{!} {
            \begin{tabular}{@{\extracolsep{5pt}}lcccc}
                \\[-1.8ex]\hline
                \hline \\[-1.8ex]
                & \multicolumn{4}{c}{\textit{Dependent variable: GDP Per Capita Growth (bps)}} \
                \cr \cline{2-5}
                \\[-1.8ex] & \multicolumn{1}{c}{Model 1 (base)} & \multicolumn{1}{c}{Model 2} & \multicolumn{1}{c}{Model 3 (Time FE)} & \multicolumn{1}{c}{Model 4 (Time + Entity FE)}  \\
                \hline \\[-1.8ex]
                 PISA Math in global P99 & & -43.491$^{}$ & -40.291$^{}$ & -126.152$^{*}$ \\
                & & (48.245) & (40.349) & (64.627) \\
                 PISA Math 99 x GDP PC & & 0.000$^{}$ & 0.000$^{}$ & 0.001$^{}$ \\
                & & (0.001) & (0.001) & (0.001) \\
                 IMO score per log population & 6.604$^{}$ & 3.629$^{}$ & 4.581$^{}$ & 20.285$^{}$ \\
                & (8.473) & (10.582) & (8.833) & (21.169) \\
                 IMO score x GDP pc & -0.000$^{}$ & 0.000$^{}$ & -0.000$^{}$ & -0.001$^{*}$ \\
                & (0.000) & (0.000) & (0.000) & (0.000) \\
                 ARWU insitutions & -588.767$^{***}$ & -538.997$^{***}$ & -580.030$^{***}$ & -764.617$^{***}$ \\
                & (135.656) & (178.746) & (149.452) & (250.804) \\
                 ARWU insitutions x GDP PC & 0.009$^{***}$ & 0.008$^{**}$ & 0.009$^{***}$ & 0.011$^{***}$ \\
                & (0.003) & (0.003) & (0.003) & (0.004) \\
                 PISA Math & & 1.150$^{}$ & 0.908$^{}$ & 0.495$^{}$ \\
                & & (0.792) & (0.732) & (1.761) \\
                 GDP per capita & -0.003$^{**}$ & -0.003$^{*}$ & -0.003$^{**}$ & 0.000$^{}$ \\
                & (0.001) & (0.002) & (0.001) & (0.005) \\
                 Primary School Completion Rate & -3.554$^{}$ & -11.911$^{***}$ & -5.387$^{}$ & -4.028$^{}$ \\
                & (3.345) & (4.157) & (3.602) & (4.947) \\
                 Lower Sec. Completion Rate & -1.475$^{}$ & -0.571$^{}$ & -1.291$^{}$ & -0.446$^{}$ \\
                & (2.945) & (3.540) & (2.948) & (4.443) \\
                 Upper Sec. Completion Rate & 5.643$^{**}$ & 6.665$^{**}$ & 5.385$^{**}$ & 3.364$^{}$ \\
                & (2.343) & (2.835) & (2.354) & (3.692) \\
                 Democracy Rating & -1.606$^{}$ & 2.270$^{}$ & -7.429$^{}$ & 36.204$^{}$ \\
                & (13.894) & (17.254) & (14.431) & (61.854) \\
                 Time Effects & Yes & No & Yes & Yes \\
                 Fixed Effects & No & No & No & Yes \\
                 Entities & 89 & 89 & 89 & 89 \\
                \hline \\[-1.8ex]
                 Observations & 440 & 440 & 440 & 440 \\
                 $R^2$ & 0.388 & 0.092 & 0.392 & 0.567 \\
                 Adjusted $R^2$ & 0.364 & 0.064 & 0.364 & 0.428 \\
                 Residual Std. Error & 360.908 (df=423) & 437.923 (df=426) & 360.976 (df=420) & 342.461 (df=332) \\
                 F Statistic & 16.727$^{***}$ (df=16; 423) & 3.314$^{***}$ (df=13; 426) & 14.230$^{***}$ (df=19; 420) & 4.066$^{***}$ (df=107; 332) \\
                \hline
                \hline \\[-1.8ex]
                \textit{Note:} & \multicolumn{4}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\
                \end{tabular}
        }
        \end{table}
\end{frame}

@@

@@

>>>>

\begin{frame}{Table of Contents}
    \tableofcontents
\end{frame}
\section{Intro}
\begin{frame}
    \frametitle{Research Question}
    \begin{block}{What is the relationship between elite students, academic institutions, and economic growth?}
        \begin{itemize}
            \item Are there outsized returns to economic growth when there are more academic elites?
            \item How can we measure academic competition and education quality beyond means?
        \end{itemize}
    \end{block}
    \begin{exampleblock}{Method}
        Through this paper, I investigate the relationship between the number of top-ranked universities, share of top math students,
        IMO scores, and GDP per capita growth to analyze these questions.
    \end{exampleblock}
\end{frame}

\section{Data}
\begin{frame}
\frametitle{Data Sources}
\begin{itemize}
    \item World Bank: World Development Indicators
    \item International Math Olympiad
    \item ARWU University Rankings
    \item PISA Math Scores
    \item Economist Democracy Scores
\end{itemize}
\end{frame}
\subsection{World Bank}
\begin{frame}{World Bank Data}
    The World Bank collects and publishes development indicators for most countries/economies in the world. In this paper, I use the following variables:

@@

\subsection{Null Imputation}
\begin{frame}{Imputing nulls}{Using XGBoost for better predictions}
    \begin{block}{Data is not missing at random}
        School completion data is only available for a maximum of 80 countries per year and has high variance in this availablility.
        This is the most limiting factor in the analysis.
    \end{block}

@@

\subsection{IMO}
\begin{frame}{IMO Scores}{The IMO is an international mathematics contest for high school students
    }
    \begin{columns}
        \column{0.5\textwidth}
            \centering
            \begin{table}
            \caption{Top 10 countries by IMO score.}
            \resizebox{\textwidth}{!}{
            \begin{tabular}{lrrr}
                \toprule
                 & IMO Score & GDPpc & GDPpc Growth \\
                Region &  &  &  \\
                \midrule
                KOR & 10.2026 & 26060.9203 & 293.8552 \\
                CHN & 9.8296 & 6424.4972 & 786.8171 \\
                USA & 9.6196 & 54346.5078 & 125.7613 \\
                RUS & 9.0800 & 10432.8366 & 277.0704 \\
                SGP & 9.0642 & 51652.4651 & 348.0543 \\
                BGR & 8.8821 & 7556.3933 & 417.7564 \\
                ROU & 8.7367 & 9432.6698 & 439.7522 \\
                HUN & 8.7153 & 13979.7009 & 262.9650 \\
                VNM & 8.5029 & 2159.2583 & 527.6404 \\
                UKR & 8.3941 & 3079.8378 & 131.9303 \\
                \bottomrule
            \end{tabular}}
        \end{table}
        \column{0.5\textwidth}
            \begin{itemize}
                \item Scores collected from 2003 to 2022
                \item Scores are transformed by $t(s) = \frac{s}{\log P}$ where $s$ is the country's raw score and $P$ is population.
                \begin{itemize}
                    \item Team size of 6 means that larger countries have an advantage due to ``genius odds''
                    \item Score is capped, so dividing by $\log P$ will correct for theoretical ceiling of performance
                \end{itemize}
            \end{itemize}
        \end{columns}
\end{frame}

\subsection{ARWU Rankings}
\begin{frame}{ARWU Rankings}{Description and Usage}
    ARWU (Academic Ranking of World Universities) is a set of university rankings based primarily on research output.
    \begin{itemize}
        \item Rankings are produced annually and are available from 2003
        \item From 2003 to 2016, 500 top universities were ranked; after 2017, 1000 were ranked.
    \end{itemize}

@@

\subsection{EIU Democracy Index}
\begin{frame}{EIU Democracy Index}
    \begin{columns}
        \column{0.4\textwidth}
        Index measuring the quality of democracy around the world published by the Economist Intelligence Unit.
        \begin{itemize}
            \item Published from 2006 every 2 years until 2010, annually afterwards (use XGBoost to impute missing)
            \item 0-10 scale, where 10 is democracy and 0 is autocracy.
        \end{itemize}

@@

\end{frame}
\subsection{Summary}
\begin{frame}{Summary Statistics}
    \begin{table}
        \caption{Summary Statistics}
        \resizebox{\linewidth}{!} {
            \begin{tabular}{lrrrrr}
                & Count & Mean & Std & Min & Max \\
                \\[-1.8ex]\hline
                \hline \\[-1.8ex]
               GDP per capita & 441.00 & 28463.99 & 25255.10 & 543.11 & 149461.79 \\
               GDP per capita growth (bps) & 440.00 & 169.06 & 452.68 & -2292.68 & 3303.05 \\
               EIU Democracy Index & 441.00 & 7.25 & 1.77 & 1.93 & 9.93 \\
               Primary Completion & 441.00 & 90.00 & 10.25 & 51.35 & 101.95 \\
               Lower Sec. Completion & 441.00 & 77.89 & 17.14 & 29.21 & 101.97 \\
               Upper Sec. Completion & 441.00 & 62.20 & 18.48 & 18.44 & 97.40 \\
               Population & 441.00 & 35318600.19 & 59445620.16 & 34000.00 & 333287557.00 \\
               ARWU per million pop ranked & 441.00 & 0.26 & 0.33 & 0.00 & 1.54 \\
               IMO score per log pop & 441.00 & 4.72 & 3.49 & 0.00 & 11.79 \\
               PISA math in global 1\% & 441.00 & 0.94 & 1.46 & 0.00 & 14.64 \\
               PISA math & 441.00 & 461.93 & 56.23 & 315.96 & 574.66 \\
               \end{tabular}
        }
    \end{table}
    \begin{table}
        \tiny
        \caption{Pre-imputation Statistics}
        \begin{tabular}{lrrrrr}
         & Count & Mean & Std & Min & Max \\
         \\[-1.8ex]\hline
         \hline \\[-1.8ex]
        EIU Democracy Index & 323.00 & 7.09 & 1.78 & 1.93 & 9.93 \\
        Primary Completion & 157.00 & 90.29 & 10.31 & 51.35 & 100.00 \\
        Lower Sec. Completion & 179.00 & 76.63 & 20.03 & 29.21 & 99.98 \\
        Upper Sec. Completion & 198.00 & 62.32 & 20.58 & 18.44 & 97.40 \\
        \end{tabular}
        \end{table}
\end{frame}

@@

\section{Empirical Results}
\begin{frame}{Model specification}
    Let elite indicators be: $math99, ARWU, IMO$.
    For the sake of concision, let $E_{i,t}$ be a 1 by 4 matrix defined as:
    \[E_{i,t} = 
    \begin{bmatrix}
        math99_{i, t} & ARWU_{i, t} & IMO_{i, t}
    \end{bmatrix}
    \] for country/region $i$ in year $t$.
    Let the coefficients of $E_{i, t}$ be a 4 by 1 matrix called $\lambda$. For control variables, $I_{i, t}$ is the matrix of variables and $\alpha$ is coefficients.
    \begin{equation}
        Y_{i, t} = \beta_0 + \lambda E_{i, t} + \delta GDPpc \times E_{i,t} + \alpha I_{i, t} + T_t + C_i + \epsilon_{i, t}
    \end{equation}
    where $T,C$ represent time and entity dummies respectively and $Y_{i,t}$ be the GDP per capita growth in basis points for a country/region $i$ and year $t$.

    Different specifications where $T_t, C_i$ are added/removed are also tested. Yearly specification without fixed effects also tested.
\end{frame}

\begin{frame}{Relationships}
    \centering
    \includegraphics[width=\textwidth]{../charts/relationships.png}
\end{frame}
\subsection{Regression tables}
\begin{frame}{PISA panel regression}{Highly dependent on model specification due to rich country bias and limited variation}
    \begin{table}[!htbp] \centering
        \resizebox{\linewidth}{!} {
            \begin{tabular}{@{\extracolsep{5pt}}lcccc}
                \\[-1.8ex]\hline
                \hline \\[-1.8ex]
                & \multicolumn{4}{c}{\textit{Dependent variable: GDP Per Capita Growth (bps)}} \
                \cr \cline{2-5}
                \\[-1.8ex] & \multicolumn{1}{c}{Model 1 (base)} & \multicolumn{1}{c}{Model 2} & \multicolumn{1}{c}{Model 3 (Time FE)} & \multicolumn{1}{c}{Model 4 (Time + Entity FE)}  \\
                \hline \\[-1.8ex]
                 PISA Math in global P99 & & -43.491$^{}$ & -40.291$^{}$ & -126.152$^{*}$ \\
                & & (48.245) & (40.349) & (64.627) \\
                 PISA Math 99 x GDP PC & & 0.000$^{}$ & 0.000$^{}$ & 0.001$^{}$ \\
                & & (0.001) & (0.001) & (0.001) \\
                 IMO score per log population & 6.604$^{}$ & 3.629$^{}$ & 4.581$^{}$ & 20.285$^{}$ \\
                & (8.473) & (10.582) & (8.833) & (21.169) \\
                 IMO score x GDP pc & -0.000$^{}$ & 0.000$^{}$ & -0.000$^{}$ & -0.001$^{*}$ \\
                & (0.000) & (0.000) & (0.000) & (0.000) \\
                 ARWU insitutions & -588.767$^{***}$ & -538.997$^{***}$ & -580.030$^{***}$ & -764.617$^{***}$ \\
                & (135.656) & (178.746) & (149.452) & (250.804) \\
                 ARWU insitutions x GDP PC & 0.009$^{***}$ & 0.008$^{**}$ & 0.009$^{***}$ & 0.011$^{***}$ \\
                & (0.003) & (0.003) & (0.003) & (0.004) \\
                 PISA Math & & 1.150$^{}$ & 0.908$^{}$ & 0.495$^{}$ \\
                & & (0.792) & (0.732) & (1.761) \\
                 GDP per capita & -0.003$^{**}$ & -0.003$^{*}$ & -0.003$^{**}$ & 0.000$^{}$ \\
                & (0.001) & (0.002) & (0.001) & (0.005) \\
                 Primary School Completion Rate & -3.554$^{}$ & -11.911$^{***}$ & -5.387$^{}$ & -4.028$^{}$ \\
                & (3.345) & (4.157) & (3.602) & (4.947) \\
                 Lower Sec. Completion Rate & -1.475$^{}$ & -0.571$^{}$ & -1.291$^{}$ & -0.446$^{}$ \\
                & (2.945) & (3.540) & (2.948) & (4.443) \\
                 Upper Sec. Completion Rate & 5.643$^{**}$ & 6.665$^{**}$ & 5.385$^{**}$ & 3.364$^{}$ \\
                & (2.343) & (2.835) & (2.354) & (3.692) \\
                 Democracy Rating & -1.606$^{}$ & 2.270$^{}$ & -7.429$^{}$ & 36.204$^{}$ \\
                & (13.894) & (17.254) & (14.431) & (61.854) \\
                 Time Effects & Yes & No & Yes & Yes \\
                 Fixed Effects & No & No & No & Yes \\
                 Entities & 89 & 89 & 89 & 89 \\
                \hline \\[-1.8ex]
                 Observations & 440 & 440 & 440 & 440 \\
                 $R^2$ & 0.388 & 0.092 & 0.392 & 0.567 \\
                 Adjusted $R^2$ & 0.364 & 0.064 & 0.364 & 0.428 \\
                 Residual Std. Error & 360.908 (df=423) & 437.923 (df=426) & 360.976 (df=420) & 342.461 (df=332) \\
                 F Statistic & 16.727$^{***}$ (df=16; 423) & 3.314$^{***}$ (df=13; 426) & 14.230$^{***}$ (df=19; 420) & 4.066$^{***}$ (df=107; 332) \\
                \hline
                \hline \\[-1.8ex]
                \textit{Note:} & \multicolumn{4}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\
                \end{tabular}
        }
        \end{table}
\end{frame}

@@

\subsection{Preliminary Interpretation}

@@

\section{Appendix}
\begin{frame}{Appendix: convergence and divergence in elite indicators}
    \centering
    \includegraphics[width=\textwidth]{../charts/std-dev.png}
\end{frame}

\begin{frame}{Appendix: elite indicators and GDP per capita}
    \centering
    \includegraphics[width=\textwidth]{../charts/relationships-gdp-pc.png}
\end{frame}
